String Constraints with Regex-Counting and String-Length Solved Efficiently

### 2023-05-09 2023-1-19/sections/introduction.tex
some syntax error in intro

In modern programming languages---such as JavaScript, Python, Java, and PHP---the string data type plays a crucial role. 
It is well-known that string manipulations are error-prone and could even give rise to security vulnerabilities (e.g.\ cross-site scripting, aka XSS). 
One powerful method for identifying such bugs in programs is \emph{symbolic execution} (possibly in combination with dynamic analysis), which analyses symbolic paths in a program by viewing them as constraints, whose feasibility is checked by constraint solvers. 
%
Symbolic execution of string manipulating programs have motivated the highly active research area of \emph{string constraint solving}, resulting in the development of numerous string solvers in the last decade or so including the state-of-the-art string constraint solvers (e.g.
Z3~\cite{Z3}, CVC4/5~\cite{cvc4}, Z3str/2/3/4~\cite{Z3-str,Z3-str2,Z3-str3,BerzishMurphy2021}, Z3str3RE~\cite{BD+23}, 
Z3-Trau~\cite{Z3-trau}, OSTRICH~\cite{CHL+19}, Slent~\cite{WC+18}, among many others.

Regular expressions (abbreviated as regex) and string-length function are widely used in string manipulating programs. According to the statistics from \cite{CS16,DCSL18,WS18}, regular expressions are used in about 30–40\% Java, JavaScript, and Python software projects. 
Moreover, the statistics from \cite{malware_detection_3_kudzu} show that string-length function occupies 78\% of the occurrences of string operations in 18 Javascript applications. 
%Operations like \verb|length| that take string inputs and return an integer frequently appear in real-world JavaScript applications (78\% of string operations in 18 applications)
As a result, most of the aforementioned string constraint solvers support both regular expressions and string-length function. Moreover, specialized algorithms have been proposed to solve efficiently the class of string constraints with regular expressions and string-length function \cite{LTR+15,BD+23}. 
%a decision procedure based on tableaux calculus was designed  in \cite{LTR+15} for the class of string constraints with regular expressions and string-length function and the decision procedure has been implemented in CVC4/5 \cite{LRT+16}.

Counting (aka repetition) is a convenient feature in regular expressions that count the number of matchings of sub-patterns. For instance, $a^{\{2, 4\}}$ specifies that $a$ occurs for at least twice and at most four times and $a^{\{2, \infty\}}$ specifies that $a$ occurs at least twice. 
Note that Kleene star and Kleene plus operator can be seen as special cases of counting operator, for instance, $a^*$ and $a^+$ are equivalent to $a^{\{0,\infty\}}$ and $a^{\{1,\infty\}}$.
Counting is also a frequently used feature of regular expressions. According to the statistics from \cite{CS16}, Kleene star/plus occur in more than 70\% of 1,204 Python projects, while the other forms of counting occur in more than 20\% of them.  Therefore, an efficient analysis of string manipulating programs requires efficient solvers for string constraints that contain at least regular expressions with counting and string-length function.

Nevertheless, the aforementioned state-of-the-art string constraint solvers still suffer from the string constraints with regex-counting and string-length, especially when the counting bounds are large. For instance, none of string solvers e.g. CVC4/5, Z3, Z3str3RE, OSTRICH is capable of solving the following constraint within 300 seconds,
%
\begin{equation}\label{eqn-running}
x \in (\Sigma \setminus a)^{\{1, 60\}} (\Sigma \setminus b)^{\{1, 60\}} (\Sigma \setminus c)^{\{0, 60\}} \wedge x \in \Sigma^* c^+ \wedge |x| > 120.
\end{equation}
Intuitively, the constraint in (\ref{eqn-running}) specifies that $x$ is a concatenation of three subwords $x_1, x_2, x_3$ where $a$ (resp. $b, c$) does not occur in $x_1$ (resp. $x_2, x_3$), moreover, $x$ ends with a nonempty sequence of $c$'s, and the length of $x$ is greater than $120$. It is easy to observe that this constraint is unsatisfiable since on the one hand, $|x| > 120$ and the counting upper bound $60$ in  both $(\Sigma \setminus a)^{\{1, 60\}}$ and $(\Sigma \setminus b)^{\{1, 60\}}$ imply that $x$ must end with some character from $\Sigma \setminus c$, that is, a character different from $c$, and on the other hand, $x \in \Sigma^*c^+$ requires that $x$ has to end with $c$.

A typical way of string constraint solvers to deal with regex-counting is to unfold regular expressions with counting into those \emph{without} counting by utilizing the concatenation operator. For instance, $a^{\{1, 4\}}$ is unfolded into $a(\varepsilon + a + aa + aaa)$. Since the unfolding incurs an exponential blow-up on the constraint sizes (where the integer constants are encoded in binary), the unfolding penalizes the performance of the solvers, especially when the bounds are large.

@@

The representation of regex-counting and string-length as registers in CEFA, combined with the aforementioned size-reduction and under-approximation techniques, entails an efficient algorithm for solving RECL constraints. We implement the algorithm on top of OSTRICH, resulting into a new version of OSTRICH, called $\ostrichrecl$ for convenience. 
%
Furthermore, we utilize a collection of benchmark suites, which comprise 49,379 RECL constraints in total, to evaluate the performance of $\ostrichrecl$. The collection of benchmark suites includes a widely-used benchmark suite for string constraints, as well as benchmark suites comprising constraints generated from regular expressions in various open source regex libraries. 
%
%Note that in the benchmark suites, we only consider the regexes that contain explicit counting, that is, regexes that contains sub-expressions of the form $e^{\{n, m\}}$, and ignore those that contain only implicit counting operators, that is, $e^*$ or $e^+$, since the state-of-the-art string solvers suffer mainly from the explicit counting operators. 
%
The experimental results show that $\ostrichrecl$ can solve 49,128 out of 49,379 instances, while the state-of-the-art best solver Z3str3RE solves only 47,407 instances, with the timeout set as 60 seconds. Moreover, $\ostrichrecl$ and Z3str3RE spend almost the same amount of time in average, that is, 2.5 seconds, in solving each instance. 
In particular, the string constraint in~(\ref{eqn-running}) can be solved by $\ostrichrecl$ within one second. 
In comparison, CVC5, Z3str3, and Z3seq solve only 44,001, 36,320, 46,813 instances respectively, with an average time 7.8, 7.7, 3.8 seconds per instance.

At last, we would like to clarify the relationship of this work with~\cite{atva2020}, where CEFA that was introduced. This work differs from \cite{atva2020} in the following aspects: In \cite{atva2020}, the regex-counting was unfolded explicitly. Moreover, in~\cite{atva2020}, the nonemptiness of CEFA was solved by utilizing the model checker nuXmv~\cite{nuxmv}, while in this work, it is solved by integrating the techniques of Parikh image computation, automata-size reduction, under approximation, and SMT solvers for LIA. The experimental results show that our approach achieves better performance than that in~\cite{atva2020}. \zhilin{this point should be supposed by experimental results.}

\paragraph*{Related work} 
We discuss more related work about regular expressions with counting and real-world regular expressions in programming languages here.  
Determinism of regular expressions with counting was investigated in~\cite{GGM12,CL15}. Counting-set automata was proposed in~\cite{redos_lenka,HS+23} to enable fast matching of a subclass of regular expressions with counting.  Moreover, a variant of nondeterministic counter automata, called bit vector automata, was proposed recently in \cite{GKM23} to enable fast matching of a more expressive class of regular expressions with counting.   Nevertheless, nonemptiness problem of these automata, which is the key for automata-based string constraint solving, was not considered in ~\cite{redos_lenka,HS+23,GKM23}.
Real-world regular expressions in programming languages include some features beyond classical regular expressions, e.g. the greedy/lazy Kleene star, capturing groups, and back references. Real-world regular expressions haven been addressed in symbolic execution of Javascript programs~\cite{LMK19} and string constraint solving \cite{CF+22}.

@@

Recently, Z3str3 has been extended to BD+23\cite{BD+23} by Berzish et al to handle string constants with regular memberships and length operation effectively. It uses automata to formalize the sematic of regular expression and applies several length-aware heuristics. BD+23 can avoid building unsolvable automaton because the length abstraction is directly derived from the regex, not the automaton. Unfortunately, the length abstraction is an over-approximation, so it can not solve string constraints (\ref{eq:BD+23_timeout}) whose regular membership predicate contains complementary.

>>>>

In modern programming languages---such as JavaScript, Python, Java, and PHP---the string data type plays a crucial role. 
It is well-known that string manipulations are error-prone and could even give rise to security vulnerabilities (e.g. \ cross-site scripting, aka XSS). 
One powerful method for identifying such bugs in programs is \emph{symbolic execution} (possibly in combination with dynamic analysis), which analyses symbolic paths in a program by viewing them as constraints, whose feasibility is checked by constraint solvers. 
%
Symbolic execution of string manipulating programs has motivated the highly active research area of \emph{string constraint solving}, resulting in the development of numerous string solvers in the last decade or so, including the state-of-the-art string constraint solvers (e.g.
Z3~\cite{Z3}, CVC4/5~\cite{cvc4}, Z3str/2/3/4~\cite{Z3-str,Z3-str2,Z3-str3,BerzishMurphy2021}, Z3str3RE~\cite{BD+23}, 
Z3-Trau~\cite{Z3-trau}, OSTRICH~\cite{CHL+19}, Slent~\cite{WC+18}, among many others.

Regular expressions (abbreviated as regex) and string-length function are widely used in string-manipulating programs. According to the statistics from \cite{CS16,DCSL18,WS18}, regular expressions are used in about 30–40\% Java, JavaScript, and Python software projects. 
Moreover, the statistics from \cite{malware_detection_3_kudzu} show that string-length function occupies 78\% of the occurrences of string operations in 18 Javascript applications. 
%Operations like \verb|length| that take string inputs and return an integer frequently appear in real-world JavaScript applications (78\% of string operations in 18 applications)
As a result, most of the aforementioned string constraint solvers support both regular expressions and string-length function. Moreover, specialized algorithms have been proposed to efficiently solve the class of string constraints with regular expressions and string-length function \cite{LTR+15,BD+23}. 
%a decision procedure based on tableaux calculus was designed in \cite{LTR+15} for the class of string constraints with regular expressions and string-length function. The decision procedure has been implemented in CVC4/5 \cite{LRT+16}.

Counting (aka repetition) is a convenient feature in regular expressions that count the number of matchings of sub-patterns. For instance, $a^{\{2, 4\}}$ specifies that $a$ occurs at least twice and at most four times, and $a^{\{2, \infty\}}$ specifies that $a$ occurs at least twice. 
Note that Kleene star and Kleene plus operator can be seen as special cases of counting operator. For instance, $a^*$ and $a^+$ are equivalent to $a^{\{0,\infty\}}$ and $a^{\{1,\infty\}}$.
Counting is also a frequently used feature of regular expressions. According to the statistics from \cite{CS16}, Kleene star/plus occur in more than 70\% of 1,204 Python projects, while the other forms of counting occur in more than 20\% of them. Therefore, an efficient analysis of string manipulating programs requires efficient solvers for string constraints that contain at least regular expressions with counting and string-length function.

Nevertheless, the aforementioned state-of-the-art string constraint solvers still suffer from the string constraints with regex-counting and string-length, especially when the counting bounds are large. For instance, none of the string solvers e.g. CVC4/5, Z3, Z3str3RE, OSTRICH is capable of solving the following constraint within 300 seconds,
%
\begin{equation}\label{eqn-running}
x \in (\Sigma \setminus a)^{\{1, 60\}} (\Sigma \setminus b)^{\{1, 60\}} (\Sigma \setminus c)^{\{0, 60\}} \wedge x \in \Sigma^* c^+ \wedge |x| > 120.
\end{equation}
Intuitively, the constraint in (\ref{eqn-running}) specifies that $x$ is a concatenation of three subwords $x_1, x_2, x_3$ where $a$ (resp. $b, c$) does not occur in $x_1$ (resp. $x_2, x_3$), moreover, $x$ ends with a nonempty sequence of $c$'s, and the length of $x$ is greater than $120$. It is easy to observe that this constraint is unsatisfiable since on the one hand, $|x| > 120$ and the counting upper bound $60$ in both $(\Sigma \setminus a)^{\{1, 60\}}$ and $(\Sigma \setminus b)^{\{1, 60\}}$ imply that $x$ must end with some character from $\Sigma \setminus c$, that is, a character different from $c$, and on the other hand, $x \in \Sigma^*c^+$ requires that $x$ has to end with $c$.

A typical way for string constraint solvers to deal with regex-counting is to unfold regular expressions with counting into those \emph{without} counting by using the concatenation operator. For instance, $a^{\{1, 4\}}$ is unfolded into $a(\varepsilon + a + aa + aaa)$. Since the unfolding incurs an exponential blow-up on the constraint sizes (where the integer constants are encoded in binary), the unfolding penalizes the performance of the solvers, especially when the bounds are large.

@@

The representation of regex-counting and string-length as registers in CEFA, combined with the aforementioned size-reduction and under-approximation techniques, entails an efficient algorithm for solving RECL constraints. We implement the algorithm on top of OSTRICH, resulting in a new version of OSTRICH, called $\ostrichrecl$ for convenience. 
%
Furthermore, we utilize a collection of benchmark suites, which comprise 49,379 RECL constraints in total, to evaluate the performance of $\ostrichrecl$. The collection of benchmark suites includes a widely-used benchmark suite for string constraints, as well as benchmark suites comprising constraints generated from regular expressions in various open source regex libraries. 
%
%Note that in the benchmark suites, we only consider the regexes that contain explicit counting, that is, regexes that contain sub-expressions of the form $e^{\{n, m\}}$, and ignore those that contain only implicit counting operators, that is, $e^*$ or $e^+$, since the state-of-the-art string solvers suffer mainly from the explicit counting operators. 
%
The experimental results show that $\ostrichrecl$ can solve 49,128 out of 49,379 instances, while the state-of-the-art best solver Z3str3RE solves only 47,407 instances, with the timeout set as 60 seconds. Moreover, $\ostrichrecl$ and Z3str3RE spend almost the same amount of time, on average, 2.5 seconds, to solve each instance. 
In particular, the string constraint in~(\ref{eqn-running}) can be solved by $\ostrichrecl$ within one second. 
In comparison, CVC5, Z3str3, and Z3seq solve only 44,001, 36,320, 46,813 instances respectively, with an average time 7.8, 7.7, 3.8 seconds per instance.

At last, we would like to clarify the relationship of this work with~\cite{atva2020}, where CEFA was introduced. This work differs from \cite{atva2020} in the following aspects: In \cite{atva2020}, the regex-counting was unfolded explicitly. Moreover, in~\cite{atva2020}, the nonemptiness of CEFA was solved by utilizing the model checker nuXmv~\cite{nuxmv}. In contrast, this work solves it by integrating the techniques of Parikh image computation, automata-size reduction, under approximation, and SMT solvers for LIA. The experimental results show that our approach achieves better performance than that in~\cite{atva2020}. \zhilin{this point should be supposed by experimental results.}

\paragraph*{Related work} 
Here, we discuss more related work about regular expressions with counting and real-world regular expressions in programming languages.  
The determinism of regular expressions with counting was investigated in~\cite{GGM12,CL15}. Counting-set automata were proposed in~\cite{redos_lenka,HS+23} to match a subclass of regular expressions with counting quickly. Moreover, a variant of nondeterministic counter automata, called bit vector automata, was proposed recently in \cite{GKM23} to enable fast matching of a more expressive class of regular expressions with counting.   Nevertheless, the nonemptiness problem of these automata was not considered in ~\cite{redos_lenka,HS+23,GKM23}, and it is unclear whether these automata models can be used for solving string constraints with regex-counting and string-length.
Real-world regular expressions in programming languages include features beyond classical regular expressions, e.g. the greedy/lazy Kleene star, capturing groups, and back references. Real-world regular expressions have been addressed in symbolic execution of Javascript programs~\cite{LMK19} and string constraint solving \cite{CF+22}.

@@

Recently, Z3str3 has been extended to BD+23\cite{BD+23} by Berzish et al to handle string constants with regular memberships and length operation effectively. It uses automata to formalize the semantic of regular expression and applies several length-aware heuristics. BD+23 can avoid building unsolvable automaton because the length abstraction is directly derived from the regex, not the automaton. Unfortunately, the length abstraction is an over-approximation, so it can not solve string constraints (\ref{eq:BD+23_timeout}) whose regular membership predicate contains complementary.
